---
title: "Objects, functions, data"
subtitle: "Tutorial 05"
author: "Psychology as a Science"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
css: css/paas_learnr.css
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = T)
question <- function(..., randomise = T, retry = T) {
  learnr::question(..., allow_retry = retry, random_answer_order = randomise)
}

```

## Intro

Welcome to the fifth PAAS tutorial.
We will pick up where we left off last time and expand on the notion of objects and functions in `R`.
Since we are learning to use `R` with the intention to analyse data, we will also talk about how `R` data classifies and structures data.

Before you proceed with this tutorial, make sure you understand everything we've covered so far.

\ 

## Objects and function calls

<div class="bquote">
Everything that exists is an object.

Everything that happens is a function call.

<div class="qauth">[John Chambers](https://en.wikipedia.org/wiki/John_Chambers_(statistician))</div>
</div>

The quote from one of the creators of `R` conveys an essential feature of the language.
It is not an exaggeration; it is actually 100% accurate: everything that exists in `R` is an object and everything we ask `R` to do is a function call.

### Function calls

A function call is simply the act of using a function.
For example, you surely know how to add two numbers together.
Addition is a function.
It takes two numbers as inputs and returns a single number as an output.
However, you don't go around constantly adding numbers, just because you know how it's done.
You evoke, or *call* the addition function stored somewhere in your brain (or your phone) only when you want to use it.
So, *addition* is a function and *adding* is a **function call**.

Function calls are represented with a set of parentheses `()` at the end of a function name, *e.g.*, `tap.on()` or `wait(600)`.

### Objects

An object in `R` is simply some sort of thing.
This may not be a very satisfying definition but bear with it for the moment.

There are two basic kinds of things/object you need to be familiar with: data and functions.
Let's revisit our kettle-filling code from last time to better explain what this means.

```{r, eval=FALSE}
kettle_capacity <- 400 # capacity of the kettle
water_needed <- 500 # water needed, kettle capacity allowing
wait_time <- 600 # how many milliseconds to wait()

tap.on() # turn the tap on

# loop while condition is TRUE
while ((get.kettle.content() < water_needed) & !is.kettle.full()) {
  wait(wait_time) # wait for wait_time ms
}

tap.off() # turn the tap off
```

\ 

#### Functions

We already talked about how `tap.on()`, `tap.of()`, `get.kettle.content()`, `is.kettle.full()`, and `wait()` are functions. They are objects that can *do* things: turn a tap on or off, check the content of the kettle, wait for a number of milliseconds.

Contrast these to all the other things we've encountered thus far: numbers and logical values (`TRUE` and `FALSE`).
These are things that contain information but don't really do anything on their own.

**We use functions to do things with data.**

You can liken this relationship to baking a cake.
Let's say I have a `make.cake()` function that contains the instructions how to mix the batter, line the cake tin, pour the batter in the tin, turn the oven on, put the tin in the oven, wait, and finally take the cake out and turn the oven off.
This function in and of itself is not enough to make a cake.
We also need the ingredients.
In this analogy, the ingredients are the data.

Take, for instance our `wait()` function.
On its own, it doesn't really do anything.
It needs some data -- namely the number of milliseconds to wait -- in order to do something: `wait(600)`.
Sure, there are other functions in the kettle-filling code that don't seem to require any data, but bear in mind that this was a simplified example and, in actual code, even these functions would need data to work with.


<div class="warn">While, technically speaking, the parentheses denote a function **call** rather than the function itself, we will write out the names of functions with them attached so that you know we're talking about functions and not about data objects</div>

OK, you might be thinking, but what are operators then?
They do things!
But they don't use parentheses?!
Good thinking!
Operators **are** functions and there is a way of using them with parentheses but, because, it makes more sense to say things like `2 + 4`, rather than `` `+`(2, 4) ``, operators have been also given a special form that allows us to use them the way we have been doing.

\ 

We will go into more detail about what functions are, how they work, and how to call them next time.
Let's focus on data now.

\ 

## Data

As we said earlier, data are objects that contain information.
By information, we mean anything that functions can operate on.
In `R` practice, it is often things like datasets, variables, test results, and so on.

### Data types

There are three types of data: logical, numeric, and character.
You've already got your hands on two of these.

- `logical` data can either be `TRUE` or `FALSE`. This is the simplest data type.
- `numeric` data are, as the name suggest, numbers. They can be positive or negative, whole or decimal numbers. Occasionally, you may see `R` refer to whole numbers as `"integer"` or `"int"` and decimal numbers as `"doulbe"` or `"dbl"`. The reason for the strange name `"double"` is that decimal numbers take up double the amount of computer memory than whole numbers. For every digit to the left of the decimal point, there must be space for one to the right of it.
- The only type we haven't worked with yet is `"character"`. These can be strings of letters, numbers, or any other characters. This is the most complex data type. The important thing to understand about character strings is that **`R` doesn't try to understand their contents and won't perform any maths with them.** Character data are **always** surrounded by quotes (`'`), or double quotes (`"`):

```{r, eval=FALSE}
3 # this is a numeric value
-5307.2455 # so is this
"3" # this is a character string
"This is also a single character string"
'sqrt(25) + 3^7' # ALSO a character string, NOT a command!
```

\

To see for yourself how character strings behave, here's a short exercise.

**First, run the code below to see that it works just fine.**

```{r char-maths, exercise=TRUE}
17 * 5 + 100
```

### \ {.empty}

**Now, put one of the numbers in quotes (single or double) and run the code again.**

### \ {.empty}

The error you just got is basically telling you that you tried to use a non-numeric value (*i.e.*, a character string) in a arithmetic function.
So remember, *no maths with strings!*

\ 

We already know that we can use logical operators to write expressions comparing numeric and logical values.
What we haven't talked about yet is that you can use the equality and non-equality operators (`==` and `!=`) to compare character strings. Try this:

```{r char-compare, exercise=TRUE}
"mind" == "blown"
```

### \ {.empty}

Since `"mind"` and `"blown"` are two different strings, the result of the comparison is `FALSE`. Now try this:

```{r char-compare-2, exercise=TRUE}
"apple" == "apple"
```

### \ {.empty}

Finally, `R` is case-sensitive. Change one of the `"apple"` strings to `"Apple"` and run the code again.

\ 

Matching strings like this is really useful for creating subsets of your data.
Like we said in the previous tutorial, if we collected some data and only wanted to look at the data for non-native English speakers, then -- assuming that the information about our participants' first language is stored in a variable called `language` -- we could use the logical expression `language != "English"`.

This kind of manipulation is the bread and butter of all data processing and analysis so make sure that logical operations with all three data types (logical, numeric, character) make sense to you.

\ 

<div class="warn">**REMEMBER** that character strings must have quotes around them. Conversely, if something has quotes around it, it is a character string, no matter what it is. If it's a numeral, or a piece of code, it doesn't matter. `R` will treat it as a meaningless string of characters.</div>

\ 

### \ {.empty}

To ask `R` what type an object is, you we can use the `class()` function. It tells you what kind of object you're looking at. Try it yourself:

```{r class-1, exercise=TRUE}
class(-4095.7201)
```
### \ {.empty}

Now, put the number above in quotes and run the code again to see that it will really become a character string.

### \ {.empty}

<div class="textbox">
Remember that **everything that exists is an object**. This is true also for the output of functions. As you can see, the name of the object class returned by the function is in quotes, which means that it is a character string. We can then say that the `class()` function returns a character string with the class of the object that we *passed to* the function.
</div>


## Data structures

`R` wouldn't be much use if we had to work with only one data point or value at a time.
Fortunately, that's not the case.
There are several ways to group data into objects, called data structures, that `R` offers.
We will look at two, namely vectors, and tibbles.

## Vectors

### \ {.empty}

The simplest way of grouping values together is to put them in a vector.
A vector is simply a line of values or *elements*.
This line can be arbitrarily long: one element, two, seven, million... even zero elements!

The basic way to create a vector is using the `c()` function. The "c" stands for "concatenate" or "combine", as in *"combine elements into a vector"*. Run the code in the exercise box below to see how the function is used.

```{r vectors-1, exercise=TRUE}
c(1, 2, 3, 4, 10, -7)
```

### \ {.empty}

The output shows you that you now have a row of numbers - a vector. 
You can now treat the vector as if it were a single element.
For example, given that our vector is composed of numbers, we can do maths with it.

See what happens if you add `+ 2` to the command and run the code:

```{r vector-arithmetic, exercise=TRUE}
c(1, 2, 3, 4, 10, -7)
```


```{r vector-arithmetic-solution}
c(1, 2, 3, 4, 10, -7) + 2
```


### \ {.empty}

Nice, the computer added 2 to every element of the vector.
You can probably see that it's useful to be able to do maths on multiple numbers at the same time with just a single command.

\ 

Non-numeric values can, of course, also be grouped into vectors. In the box below, create a vector with the names of several of your friends or family members.

```{r char-vector, exercise=TRUE}

```

<div id="char-vector-hint">
Remember:

- character strings must be `"in quotes"`
- individual values in the `c()` function must be separated with commas
</div>

You can put as many elements into the `c()` function as you like but you **must separate them with commas**.

### Data types in vectors

Vectors can only ever contain data of the same type. See what happens, when you turn any one of the numbers in the vector below into a character string by putting it in quotes.

```{r vec-class, exercise=TRUE}
c(1, -6, 3,64, -432, 2, 45)
```


### \ {.empty}

As you can see, all the values in the vector got converted to character strings. That is because, if you try to put different data types in a single vector, they will get converted to the most complex type. Character vectors are more complex than double (decimal) numbers, which are more complex than integers, which, in turn, are more complex than logical values.

\ 

One last thing about vectors.
Remember how we told you not to worry about the `[1]` that gets printed out everytime we run a command (see the output of any exercise)?
Let's talk about it now.

Before we do, though, run the following command to create a vector of all integers between 1 and 100.

```{r vector-2, exercise=TRUE}
1:100
```

### \ {.empty}

Now that we have a longer vector, you can see that there are several of these brackets, one at the beginning of every line.
This is just `R`'s way of helping you get a feel for how many elements there are in your vector.
The number in the brackets tells you the position of the elements next to it in the voctor.
`[1]` means the next element is the first one in the vector, `[42]` means it's the 42^nd^, and so on...

\ 

The `[1]`s we saw next to every piece of output we have encountered thus far suggest that all the output were vectors!
This is true.
Vectors are the most basic data structures and so even when `R` is manipulating a single value, it treats it as a vector that contains one element.

Remember the `calss()` function?

```{r, class-fun-output}
class(1:100)
```

When you run the code above (go on, do it!), the function will return `[1] "integer"`, because the vector `1:100` (integers between 1 to 100) only contains whole numbers. But as you can also see that what the function returned was **a character vector of length 1**.

<div class="textbox">
When learning `R`, it's really important to get used to thinking about function calls in terms of their output. The code `class(1:100)` is, after all just a linguistic representation of what the command does. And what it does is return a character vector of length one that contains information about the data type of the vector `1:100`.

Thinking about code in terms of what it does is very much like normal language. If somone tells you "Pass the salt, please", you don't start analysing the sentence, thinking "'pass' is the second person imperative of the verb that means to transport or transfer, 'the' is the definite article..." No, you extract the meaning of the sentence and, hopefully, pass the salt.
</div>


## Tibbles

Vectors are certainly very useful data structures but when analysing collected data, we often want to organise them in a rectangualr structure with rows and columns.
Think spreadsheets (Excel, Google sheets, *etc*).
This way we can store data from different participants in different rows of the spreadsheet and have each column represent a different variable, for example participants' age, eye colour, or number of instagram followers.

Tibbles offer precisely this kind of structure. This is what a nicely-formatted tibble looks like:

```{r, echo=FALSE}
set.seed(6669)
library(tibble)
df <- tibble(participant_id = replicate(10, paste(sample(LETTERS, 5, replace = T), collapse = "")),
             age = sample(18:58, 10, replace = T),
             eye_colour = sample(c("blue", "brown", "green"), 10, replace = T, prob = c(35, 55, 10)),
             ig_followers = rlnorm(10, meanlog = 7))

df
```

And this is what the same tibble looks like in default raw `R` output, without the fancy formatting:

```{r, echo=F, paged.print=FALSE}
df
```

\ 

As you can see, a tibble looks very much like a spreadsheet: it is a rectangular arrangement of data with cells organised into rows and columns. This means that **all rows must contain the same number of columns and all columns must contain the same number of rows**.

Also notice that, unlike vectors, **tibbles can contain data of various types**. In the example above, the first and third columns (`participant_id` and `eye_colour`, respectively) are character (`<chr>`) and the second and fourth columns are numeric (one is integer, `<int>`, and the other is double, `<dbl>`). However, because **each column of a tibble is a vector**, each column can only contain data of one type!

Finally, the columns of a tibble are always named. In our example, the names of the columns are `participant_id`, `age`, `eye_colour`, and `ig_followers`.

\ 

We will learn how to work with vectors and tibbles in future tutorials but, for now, this is all you need to know about them.

\ 

## Storing objects

### \ {.empty}

Being able to create vectors and organise your data in tibbles is great but it's not all that much use without understanding how to store the objects you create in the computer's memory.

Let's go back to a very basic command. In the box below, type in a number -- any number -- and run the code.

```{r assign-1, exercise=TRUE}

```


### \ {.empty}

Let's break down what just happened.

You wrote a command that tells the computer to create a numeric vector one element long.
This command produces some sort of output, in this case your numeric vector of length one.
The output of the command needs to "go" somewhere.
By default, it gets directed to the console (these exercise boxes are essentially virtual `R` consoles).
After the computer has "printed" the output of the command into the console, it forgets everything about it.
As far as your computer is concerned, nothing ever happened.

To make the computer remember the output of the command, we can direct the output into computer's memory, rather than into the console.
We can do this by assigning the output to a name using the assignment operator, `<-`, we saw in the previous tutorial.

It's really simple! This is the code to store *a numeric vector of length one containing the number 1526* in the memory under the name `my_num`. All you need to do now is to run it.

```{r assign-2, exercise=TRUE}
my_num <- 1526
```

### \ {.empty}

There are only a handful of rules you need to follow to store the output of a command in the memory:

- The name must be valid
  - No white spaces allowed (hence the `_`)
  - Starts with a letter, not a number of a special character.
  - The name is not a character string so **it must not be in quotes**.
- No white space allowed between the `<` and `-` in the assignment operator.
- The command to the right of the `<-` must be a valid command so watch out for errors.

<div class="warn">
Notice that when you run the code above, the number does not get printed under the box in the same way it did in the previous exercise! This is because the output of the command got directed into the memory under the name `my_num` instead of the console.

Remember, output of a command **can only ever go to one place at a time!** You can send it to the memory, write it to a file on your computer's disk, or print it on the screen.
</div>

To recall the object from the memory, simply type its name (again, never in quotes!) and run the command:

```{r assign-3, exercise=TRUE}

```

```{r assign-3-setup}
my_num <- 1526
```

```{r assign-3-solution}
my_num
```


### \ {.empty}

And there you have it, as good as new!

What `R` now showed you was just a copy of the content of `my_num`. The actual object is still stored in the memory and can be recalled again and again until you quit or restart `R` (or this tutorial).

\ 

The name `my_num` is now just a representation of the object stored under that name. This means that you can use `my_num` in the same way you would use the number 1526. For example, you can do maths with it.

Multiply `my_num` by 4.

```{r assign-4, exercise=TRUE}

```

```{r assign-4-setup}
my_num <- 1526
```

```{r assign-4-solution}
my_num * 4
```


### \ {.empty}

As you can imagine it's very useful to be able to use names as stand-in for objects, especially objects that contain a lot of data.

There's a little catch to storing objects though. Once they are in memory, stored under a name, that's it. They can be recalled, they can be removed from the memory if you wish to do it, but they cannot be changed. At least not *technically speaking*.

Let's demonstrate this on an example. Say we create a vector of character strings representing the names of the methods modules and save it under the name `modules`:

```{r}
modules <- c("PAAS", "Analysing Data", "Discovering Statistics")
```


Here they are:

```{r}
modules
```

\ 

Now, it turned out we omitted one of the modules, Quantitative and Qualitative Methods. To add an element to an existing vector, we can use the `c()` function and the vector's name:

```{r}
# combine whatever is in modules with the string "QQM" into a single vector
c(modules, "QQM")
```

The output now shows `"QQM"` added to the other module names.


### \ {.empty}

Remember, however, two things:

1. The output of a command can only ever go to one place at a time.
2. When you print out the contents of a name, you only get a **copy of the contents**

These two principles considered, it should not be too surprising to see that, if we now print out the contents of `modules`...

```{r}
modules
```

...they will be unchanged: There's no `"QQM"` at the end!

So remember, you can't *really* change an object.


### \ {.empty}

This may seem like a major drawback of `R` until you realise that, even though you cannot change an object you can **reassign** different objects to an already existing name, thereby effectively replacing the original content with a new one.

So, to "add" `"QQM"` to `modules`, we have to assign the output of the command above to the name `modules`:

```{r}
modules <- c(modules, "QQM")
```

By reassigning to the name `modules`, the computer has now forgotten what was stored under the name before and is storing only the new object:

```{r}
modules
```

\ 

<div class="warn">
Assignment and reassignment of command output to names may be a little counter-intuitive at first but understanding it is really crucial for working with `R` so make sure you wrap your head around it before you move on!
</div>


## Summary

In this tutorial, you learnt that

- everything that exists is an object
- and everything that happens is a function call
- a function call is the action of using (calling) a function
- there are two kinds of objects:
  - functions
  - data structures
- data can be one of three types:
  - logical; `TRUE` or `FALSE` - the simplest type
  - numeric; integer or double (decimal)
  - character; always in "quotes" - the most complex type

You also found out about two kinds of data structures

- vectors
  - ordered chains of values/elements
  - can be arbitrarily long
  - can only ever contain one type of data
  - easily created with the `c()` function
- tibbles
  - like spreadsheets
  - rectangular arrangements with rows and columns
  - always have named columns
  - each column is a vector, so all rules for vectors apply
  - different columns can contain different types of data

Finally, you learnt how to store objects in memory using the assignment operator `<-`

- the syntax is:
  
    ```{r, eval=FALSE}
    valid_name_to_store_object_under <- output.of(function, call)
    ```
    
- names are never in "quotes"!
- commands and names are just symbolic representations of the objects they create/contain and can be treated as such
- output of a command can only ever go in one place at a time: file, memory, or screen
- technically, you cannot change an object stored in the memory
- you can, however, reassign a different object to the same name:
  
    ```{r, eval=FALSE}
    name <- some.function.of(name)
    ```

\ 

<div class="warn">
All of the above is extra important to know, if you ever want to learn `R`. If you learn and understand these principles, working with `R` is going to be rather easy!
</div>

\ 
  
**And as always, your last task for this tutorial is to create a Week_05 RStudio project in the paas folder.**

\ 

<big>Well done!</big>

\ 
